Community biomass and species richness temporal trends reveal contrasting assembly and disassembly processes in natural food-webs

### 2024-08-02 paper/manuscript.Rmd
Rewrite discussion

# Abstract (242 / 250 words max)

Understanding how biodiversity changes is a central question in ecology, but a
challenging one because of the multidimensionality of biodiversity changes.
Although the structure of ecological networks is a crucial determinant of
community dynamics, this dimension of biodiversity changes have been greatly
overlooked. One main reason is the scarce availability of the temporal series of
ecological networks. Our knowledge of the relationships among structural
properties of ecological networks comes mainly from spatial comparison, assuming
a space-for-time equivalence, which have rarely tested. We questioned the
relationships between the temporal changes of food-web structure and the ones of
community biomass and species richness, and assessed if they differed from the
ones estimated in space. We inferred the structure of using food diet and body
size data from fish monitoring in France. We estimated the temporal trends of
food-web structure in 400 sites using a hierarchical Bayesian model and then
assessed the relationships among them using structural equation models. We found
that temporal trends of connectance and average trophic level are positively
related to ones of community biomass but surprisingly poorly related to the
temporal trends of species richness, which resonates with the energetic paradigm
of food-webs. We further found general support for the space-for-time hypothesis
for community biomass, but not for species richness. Our study
sheds a new light on the dynamics of food-web assemblage in natural settings and
how the structure of ecological networks changes according to the changes of
other biodiversity facets.

@@

@@

```{r, results=FALSE}
tar_load(glob_tps_trends_decade_no_drivers)
glob_tps_trends_decade_no_drivers
gbm <- get_effect_ci(resp = "log_bm_std", ci_lvl = "level:0.80", p = TRUE)
gr <- get_effect_ci(resp = "log_rich_std", ci_lvl = "level:0.80", p = TRUE)
gt <- get_effect_ci(resp = "w_trph_lvl_avg", ci_lvl = "level:0.80", p = FALSE)
gct <- get_effect_ci(resp = "ct_ff", ci_lvl = "level:0.80", p = FALSE)
```

```{r}
tar_load(c(r2_inla_tab, r2_inla, r2_sem))
r2_inla_report <- r2_inla %>%
  select(response, r2_mvp_marg95hpdci, r2_mvp_cond95hpdci)
r2ic <- map(
  setNames(r2_inla_report$response, r2_inla_report$response),
  ~r2_inla_report[r2_inla_report$response == .x, ]$r2_mvp_cond95hpdci
)
```

Using a Bayesian hierarchical model to estimate conjointly the temporal trends of
food-web structure at all sites, we found weak evidence of a global decrease in
community biomass (CI80%:`r gbm` by decade; median [Credible Interval], Fig. S1) but of a
global increase in species richness (CI80%: `r gr` by decade) across the `r nb_st` sites. 
We did not find evidence of a global trends in connectance and average trophic
level (CI80%: `r gct` and `r gt` respectively). The hierarchical model captured
well the variance associated to the spatial structure of the data with
conditional $R^2$ going from  `r r2ic$log_bm_std` (mean [CI95%]) for community
biomass to `r r2ic$log_rich_std` for species richness (Table S1).

@@

```{r}
tar_load(semeff_tot)
semeff_tot <- semeff_tot %>%
  mutate(across(c(response, predictor), ~str_remove(.x, "mean_")))

rdtps <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std")
    )
)
rdsp <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std", model = "spatial"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std", model = "spatial")
    )
)
```

```{r}
ridtps <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    r = get_semeff_ci(resp = .x, term = "log_rich_std", type = "indirect")
    )
)
ridsp <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    r = get_semeff_ci(resp = .x, term = "log_rich_std",
      model = "spatial", type = "indirect")
    )
)
rttps <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std", type = "total"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std", type = "total")
    )
)
rtsp <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std",
      model = "spatial", type = "total"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std",
      model = "spatial", type = "total")
    )
)
```

Using structural equation models to disentangle the direct and indirect
relationships among food-web structure descriptors, we found a strong positive
relationship between the temporal trends of community biomass and of
species richness ($r_\delta =$ `r rdtps$bm$r`, Fig. \@ref(fig:sem) a and b,
$r_\delta$: standardized slope coefficient, [95% CI]), meaning that both
temporal trends had the same direction in average. At the sites where species
richness decreased over time, community biomass also decreased over time. On the
other hand, at the sites where species richness increased over time, community
biomass also increased over time.

The temporal trends of connectance were positively linked with the ones of
community biomass ($r_\delta =$ `r rdtps$ct$bm`), also meaning that both
temporal trends had the same direction in average. That is, the sites where
food-web connectance decreased over time also had community biomass that
decreased over time and reversely. It suggests that communities that decreased
in biomass over time lose highly connected trophic species over time, and
reversely that communities gaining biomass over time also gain highly connected
trophic species over time. We found a negative relationship between the temporal
trends of connectance and the ones of species richness
($r_\delta =$ `r rdtps$ct$r`, Fig. \@ref(fig:sem)d). We further found the
species richness had an indirect positive effect on connectance, through
community biomass ($r_\delta =$ `r ridtps$ct$r`). It resulted in a relatively
weak total negative effect of species richness on connectance
($r_\delta =$ `r rttps$ct$r`), that we also found in random extinction
simulation (Fig. S2).

The temporal trends of average trophic level had a strong positive relationship
with the ones of community biomass ($r_\delta =$ `r rdtps$tlvl$bm`, Fig.
\@ref(fig:sem)e), meaning that sites where community biomass decreased over
time, average trophic level decreased over time and reversely. It means that in
communities losing biomass over time, the average biomass of the food-web moved
down to lower trophic level and vice versa. However, we found that the temporal
trends of average trophic level had no direct relationship the ones of species
richness ($r_\delta =$ `r rdtps$tlvl$r`, Fig. \@ref(fig:sem)f). It suggests that
in communities either losing or gaining species richness over time, the location
of average biomass along trophic level does not significantly change according
to species richness. However, the temporal trends of species richness had an
indirect positive relationship with the temporal trends of average trophic level
through community biomass ($r_\delta =$ `r ridtps$tlvl$r`), resulting in a total
positive relationship ($r_\delta =$ `r rttps$tlvl$r`).

@@

As for the temporal analysis, we found a positive relationship between
median community biomass and median species richness along the spatial gradient,
but half the strength as the temporal one ($r_\delta =$ `r rdsp$bm$r`, Fig.
\@ref(fig:std)). We further found a positive relationship between median
connectance and median community biomass ($r_\delta =$ `r rdsp$ct$bm`), as in
temporal trends analysis. However, we found a negative relationship between
median connectance and median species richness, but
three times stronger than in the temporal analysis ($r_\delta =$ `r rdsp$ct$r`,
Fig. \@ref(fig:std)). We also found a positive indirect effect of median
richness on median connectance through median community biomass but a weak one
($r_\delta =$ `r ridsp$ct$r`), resulting in a strong negative total effect of
median richness on median connectance ($r_\delta =$ `r rtsp$ct$r`).

We found a positive relationship between median average trophic level and median
community biomass, but half the strength as in the temporal analysis ($r_\delta
=$ `r rdsp$tlvl$bm`, Fig. \@ref(fig:std)). However, we found a strong negative
relationship between median average trophic level and median richness ($r_\delta
=$ `r rdsp$tlvl$r`), in contrast with findings in temporal analysis. We also
found a positive indirect relationship between average trophic level and  median
richness through median community biomass (`r ridsp$tlvl$r`), resulting in a
strong total negative effect of median richness on median average trophic level
(`r rtsp$tlvl$r`).

@@

We found that species richness scaled positively with community biomass, both in
spatial and temporal analysis. This result is in accordance with previous
analysis across space in stream communities [@woods_testing_2020;
@danet_species_2021] but also in grasslands, salt marches and forests
[@hector_plant_1999; @tilman_biodiversity_2014; @wu_relationship_2015;
@li_relationship_2018; but see @grace_integrative_2016]. Therefore, our results
suggest that Biodiversity-Ecosystem functioning prediction also hold in the
temporal dynamics of food-webs. The positive relationship between species
richness and total biomass might be due to positive complementarity in resource
use [@carey_determining_2011], which is held by niche differentiation
[@loreau_biodiversity_1998; @loreau_partitioning_2001] or to the higher
probability of getting high biomass species with higher species
richness [@loreau_biodiversity_2001].

Our results suggest that the disturbances that lead to decline or increase of
community biomass over time are concomitant to the decline and increase
respectively of trophic species at high trophic level. We found that both
connectance and average trophic level had a positive relationship with community
biomass over time, meaning that connectance and average trophic level increase
over time when community biomass increases over time as well, and likewise that
connectance and average trophic level decrease over time when community biomass
decreases over time. We also found that the temporal trends of connectance and
average trophic level were strongly correlated (Fig. S3). Likewise, we found
that the temporal trends of community biomass were positively associated to the
temporal trends of both number and proportion of piscivorous species
present (Fig. S3). This result is in accordance with the
energetic paradigm of food-webs which proposes that food-webs are primarily
constrained by fluxes of matter and energy [@lindeman_trophic-dynamic_1942],
which suggest that productivity constraints food chain length by energy
limitation [@jenkins_productivity_1992; @takimoto_ecosystem_2008]. Higher
trophic levels have been shown to be more sensible to ecosystem size
[@takimoto_ecosystem_2008; @post_ecosystem_2000], to resource availability and
disturbance in small systems [@post_long_2002] and particularly to disturbance
in hydrologic regimes in streams [@sabo_role_2010; @mchugh_dual_2010;
@jake_vander_zanden_global_2007]. In the same line, the metabolic theory of
ecology also predicts that top predators are highly constrained by energy
limitation [@arim_relationship_2007]. However, this interpretation holds if
the temporal trends of community biomass are related to the quantity of resource
available. Another explanation can be that top trophic level species are often
the ones that are targeted, harvested and those that are artificially introduced
for recreational fisheries [@reid_emerging_2019]. Further investigations on the
drivers of community biomass decline and increase over time would be needed to
further conclude.

In contrast, we found that the temporal trends of connectance was negatively
related to the temporal trends of species richness, in accordance with the
strong negative relationships that we found in the present study and from
previous results over spatial gradient [@dunne_network_2006;
@danet_species_2021]. Furthermore, we found that temporal trends of the
proportion of piscivorous species was negatively associated with the temporal
trends of species richness (Fig. S3), meaning that communities that increase in
richness over time have an increase in the proportion of non-piscivorous species over time
and vice versa. This result is in accordance with a meta-analysis showing the
number of species by trophic level decrease toward higher trophic level
[@turney_pyramids_2016]. It suggests that the temporal increase of species
richness found in our dataset is associated with a decrease of connectance
because species richness increases are biased toward non-piscivorous species
which have no piscivory links. We also found that the changes of average trophic
level over time were not related to the changes of species richness over time.
All together, an explanation can be that species richness changes over time,
without changes in community biomass, are more linked to non piscivorous species
having lower trophic level and relatively low biomass, which results in negative
relationships with temporal trends in connectance and no relationship with
average trophic level.

Our findings further suggest that species richness and community biomass have
antagonistic effects on connectance, respectively negative and positive. As we
found a strong positive relationship between the temporal trends of species
richness and community biomass, it results that species richness temporal trends
had an indirect positive effect on the temporal trends of both connectance and
average trophic level via community biomass. Then,  The temporal trends of
species richness had then a total effect on the temporal trends of connectance
and average trophic level which was twice lower than the one of the temporal
trends of community biomass. It suggests that the temporal assembly and
disassembly of food-webs are more sensitive to variations in community biomass
than in species richness, in other words that food-web structure is more robust
to changes in species richness than in biomass over time. Our study then
suggests that the changes of food-web structure over time related to changes in
species richness could be actually mediated by changes in community biomass.

As studies reported community biomass or species richness temporal trends
[@hallmann_more_2017; @ceballos_biological_2017; @dornelas_assemblage_2014;
@blowes_geography_2019], we document that temporal trends of species richness
and community biomass might have different consequences on the structure of the
community. In particular, the decrease of the average trophic level associated
to the loss of community biomass could have cascading effect on ecosystem
functioning as the highest predators are predicted to increase productivity and
stabilize communities [@wang_biodiversity_2018; @shanafelt_stability_2018;
@danet_species_2021]. In contrast, the loss of species richness might be linked
to loss of redundancy in the food-web, highlighting that those communities may
become more at risk regarding future perturbations [@yachi_biodiversity_1999].
Interestingly, the hypothesis that temporal trends in community biomass is
related to species of higher trophic level in our system has a strong reasoning
with the effect of human fisheries in marine systems which affect primarily fish
species at top trophic levels [@sprules_surfing_2016].

We found evidence that space-for-time substitution is valid for community
biomass but not for species richness. More specifically, we found a positive
relationship between community biomass and species richness both in space and
time. We also found that connectance was positively linked to community biomass
both over space and time, it resonates with theoretical studies showing that
higher food-web connectance can result in higher community biomass
[@thebault_food-web_2003], or lower [@poisot_trophic_2013] over space,
highlighting that the relation between connectance and community biomass may be
dependent on how food-web structure change with community biomass over space and
time. Likewise, we found that the average trophic level had a positive
relationship with community biomass over space and time, in accordance with a
theoretical study predicting an increase of community biomass
linked to the presence of higher predators [@wang_biodiversity_2018].
However, while we found a total negative effect between connectance and species
richness in space, in accordance with previous findings over space in
empirical food-webs [@dunne_network_2006; @danet_species_2021], we found a null
total effect in time. The result is even more contrasting for average trophic
level, where we found a total positive effect of species richness in time but a
negative one in space. This result is also in line with other studies which
found a mismatch between spatial and temporal patterns [@adler_contrasting_2007;
@johnson_testing_2008]. This discrepancy in space and time might be due to the
strong assumption that variations in space reflect differences in assembly
stages of ecological communities and then mirror their temporal
assembly/disassembly trajectories. Temporal variation of species richness might
relate to transitory dynamics while spatial gradients reflect aggregated
dynamics over a longer time scale including a far number of processes including
high species turnover along spatial gradients. Our results then suggest that food-web
structure comparison along space gradient might be a reasonable surrogate for
time sequence, it is probably not the case for species richness spatial
comparison.

@@

Our study suggests that species richness and community biomass temporal trends
in natural fish communities relate on different ecological processes. Biomass
changes in time are related to variations of food-web structure at the
higher trophic levels while changes in species richness are weakly related to
changes in food-web structure, in accordance with the energetic
paradigm of food-webs. Our results further highlight that the effects of species
richness changes on food-web structure are strongly mediated by antagonistic
effects of community biomass. We also find evidences which suggest that
space-for-time substitution may be relevant to predict the effects of community
biomass losses or gains on food-web structure but also that space-for-time
substitution might not work for species richness. Our study calls for further
investigations to support the generality of assembly rules in food-webs,
according to changes in community biomass and species richness. They are crucial
to better understand how biodiversity changes will affect community dynamics
and ecosystem functioning.

>>>>

# Abstract (240 / 250 words max)

Understanding food-web assembly processes is central to biological conservation
and maintenance of ecosystem functions. In particular, temporal changes in
community biomass and species richness are expected to significantly impact
food-web structure but empirical assessments are lacking so far. Here, we
investigate how temporal trends in species richness and community biomass affect
temporal trends in food-web structure. To do so, we used more than 6000 sampled
riverine fish community in 400 sites from french monitoring data. We inferred
food-web structure for each sampling event, using diet data and body size. Using
structural equation models, we estimated the effects of temporal changes in
community biomass and species richness on temporal changes in food-web
structure. We found that a positive effect of temporal trends in community
biomass on the temporal trends of connectance and average trophic level, a
result also found using spatial variations and a bioenergetic model. Temporal
trends in species richness had a positive effect on average trophic level, but
no effects on connectance. In space and in simulations, species richness had in
contrast a negative effect on average trophic level and connectance. In sum, our
study highlight that local temporal changes in biomass and species richness have
contrasting effects on food-web structure. It further reveals that temporal
changes and spatial variations in biomass lead to coherent food-web assembly
patterns, in contrast to species richness. Our study sheds a new light on
food-web assembly in natural settings.

@@

*Introduce hypothesis and the food-web metrics (connectance and average trophic
level)*

@@

```{r, results=FALSE}
tar_load(glob_tps_trends_decade_no_drivers)
gbm <- get_effect_ci(resp = "log_bm_std", ci_lvl = "level:0.80", p = TRUE)
gr <- get_effect_ci(resp = "log_rich_std", ci_lvl = "level:0.80", p = TRUE)
gt <- get_effect_ci(resp = "w_trph_lvl_avg", ci_lvl = "level:0.80", p = FALSE)
gct <- get_effect_ci(resp = "ct_ff", ci_lvl = "level:0.80", p = FALSE)
```

```{r}
tar_load(c(r2_inla_tab, r2_inla, r2_sem))
r2_inla_report <- r2_inla %>%
  select(response, r2_mvp_marg95hpdci, r2_mvp_cond95hpdci, r2_mvp_cond_mean)
r2cond <- map(
  setNames(r2_inla_report$response, r2_inla_report$response),
  ~round(r2_inla_report[r2_inla_report$response == .x, ]$r2_mvp_cond_mean, 2)
)
r2ic <- map(
  setNames(r2_inla_report$response, r2_inla_report$response),
  ~r2_inla_report[r2_inla_report$response == .x, ]$r2_mvp_cond95hpdci
)
```

```{r}
dfv <- function(
  x = summary_pred_station_tps10,
  r = "ct_ff",
  v = "min",
  d = 4
) {
  out <- x[x$response == r, ][[v]]
  format(round(out, d), nsmall = d)
}
grbmn <- dfv(r = "log_bm_std", v = "min", d = 0)
grbmx <- dfv(r = "log_bm_std", v = "max", d = 0)
grrhn <- dfv(r = "log_rich_std", v = "min", d = 0)
grrhx <- dfv(r = "log_rich_std", v = "max", d = 0)
grctn <- dfv(r = "ct_ff", v = "min", d = 2)
grctx <- dfv(r = "ct_ff", v = "max", d = 2)
grtln <- dfv(r = "w_trph_lvl_avg", v = "min", d = 2)
grtlx <- dfv(r = "w_trph_lvl_avg", v = "max", d = 2)
```

Using a Bayesian hierarchical model to jointly estimate the temporal trends of
food-web structure at all sites, we found weak evidence of an overall decrease
in community biomass (CI80%:`r gbm` by decade; median [Credible Interval], Fig.
S1) but a weak evidence of an overall temporal increase in species richness
(CI80%: `r gr` by decade) across the `r nb_st` sites. We did not find evidence
of average temporal trends in connectance and average trophic level (CI80%:
`r gct` and `r gt` respectively). However, we found a great heterogeneity in the
temporal trends estimated at the site level (community biomass: from `r grbmn`%
to `r grbmx`% per decade, species richness: `r grrhn` to +`r grrhx`%, average
trophic level: `r grtln` to +`r grtlx`, connectance: `r grctn` to +`r grctx`,
Fig. S1) which was well captured by the hierarchical model with conditional R2
going from `r r2cond$log_bm_std` for community biomass to `r r2cond$log_rich_std`
for species richness (Table S1).

@@

```{r}
tar_load(c(semeff_tot, semeff_tot_ok))
semeff_tot <- semeff_tot %>%
  mutate(across(c(response, predictor), ~str_remove(.x, "mean_")))
unique(semeff_tot_ok$predictor)

rdtps <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std")
    )
)
rdsp <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std", model = "spatial"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std", model = "spatial")
    )
)
rdm <- map(
  setNames(
    c("Connectance", "Avg trophic level", "Biomass"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    bm = get_semeff_ci(x = semeff_tot_ok, type = "Direct", resp = .x, term = "Biomass", model = "Theoretical model"),
    r = get_semeff_ci(x = semeff_tot_ok, type = "Direct", resp = .x, term = "Species richness", model = "Theoretical model")
    )
)
```

```{r}
ridtps <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    r = get_semeff_ci(resp = .x, term = "log_rich_std", type = "indirect")
    )
)
ridsp <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    r = get_semeff_ci(resp = .x, term = "log_rich_std",
      model = "spatial", type = "indirect")
    )
)
ridm <- map(
  setNames(
    c("Connectance", "Avg trophic level", "Biomass"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    bm = get_semeff_ci(x = semeff_tot_ok, type = "Indirect", resp = .x, term = "Biomass", model = "Theoretical model"),
    r = get_semeff_ci(x = semeff_tot_ok, type = "Indirect", resp = .x, term = "Species richness", model = "Theoretical model")
    )
)
rttps <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std", type = "total"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std", type = "total")
    )
)
rtsp <- map(
  setNames(
    c("ct_ff", "w_trph_lvl_avg", "log_bm_std"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std",
      model = "spatial", type = "total"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std",
      model = "spatial", type = "total")
    )
)
rtm <- map(
  setNames(
    c("Connectance", "Avg trophic level", "Biomass"),
    c("ct", "tlvl", "bm")
    ),
  ~ list(
    bm = get_semeff_ci(x = semeff_tot_ok, type = "Total", resp = .x, term = "Biomass", model = "Theoretical model"),
    r = get_semeff_ci(x = semeff_tot_ok, type = "Total", resp = .x, term = "Species richness", model = "Theoretical model")
    )
)
```

Using structural equation models to disentangle the direct and indirect effects
of temporal trends in fish species richness and community biomass on temporal
trends in food-web structure, we found a strong positive effect of the temporal
trends in species richness on community biomass ($r_\delta =$ `r rdtps$bm$r`,
Fig. \@ref(fig:sem) a and b, $r_\delta$: standardized slope coefficient, [95%
CI]), meaning that both temporal trends had the same direction in average. In
other word, food-webs that have experienced a decrease in species richness over
time have decreased in community biomass, and conversely, those that increased
in species richness have increased in community biomass.
Temporal trends in community biomass had in turn a positive effect on connectance and
average trophic level (resp. $r_\delta =$ `r rdtps$ct$bm` and `r rdtps$tlvl$bm`,
Fig. \@ref(fig:sem)a, c, e), meaning that food-web whose community biomass
decreased over time have decreased in connectance and average trophic level, and
conversely, those that have experienced an increase in biomass have increased in
connectance and average trophic level. We further found that the temporal
trends of community biomass were positively associated to the temporal trends of
both number and proportion of piscivorous species present (Fig. S3a), meaning
biomass trends are linked to the presence of high trophic level trophic
species.

We further tested if we find the same patterns using spatial variations in
food-web structure rather than temporal changes over time, i.e. if the
space-for-time hypothesis hold. To do so, we used the same dataset and same
structural equation model but with the median food-web structure per site
instead of the temporal trends. We additionally performed theoretical
simulations with a bioenergetic food-web model to further get theoretical
expectations about how food-web structure changes according to community biomass and
species richness variations (see Methods). As in temporal data, species richness
had a positive effect on community biomass both in empirical data and in
theoretical simulations (resp. $r_\delta =$ `r rdsp$bm$r` and `r rdm$bm$r`, Fig.
\@ref(fig:std)). In both empirical data and theoretical simulations, community
biomass had a positive effect on connectance (resp. $r_\delta =$ `r rdsp$ct$bm`
and `r rdm$ct$bm`, Fig. \@ref(fig:std)) and average trophic level (resp.
$r_\delta =$ `r rdsp$tlvl$bm` and `r rdm$tlvl$bm`), as in temporal analysis.

The effects of species richness were more complex, because species richness had
indirect effects on food-web structure through community biomass. Temporal
trends in species richness had a direct negative effect on connectance
($r_\delta =$ `r rdtps$ct$r`, Fig. \@ref(fig:sem)d), but had also a positive
indirect effect through community biomass ($r_\delta =$ `r ridtps$ct$r`), so
that species richness temporal trends had no effect in total on temporal in
connectance ($r_\delta =$ `r rttps$ct$r`, Fig. \@ref(fig:sem)d).
In contrast, temporal trends in species had no direct effects on the temporal
trends in average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig.
\@ref(fig:sem)f), however it had a positive indirect effect through community
biomass ($r_\delta =$ `r ridtps$tlvl$r`). It means food-webs that have
experienced a decrease in species richness over time have decreased in average
trophic level, and conversely, those that increased in species richness have
increased in average trophic level, through community biomass. Furthermore, we
found that temporal trends of the proportion of piscivorous species was
negatively associated with the temporal trends of species richness (Fig. S3),
meaning that communities that increase in richness over time have an increase in
the proportion of non-piscivorous species over time and vice versa.

The effects of spatial variations in species richness on food-web structure
contrasts with the ones found with temporal trends. We found a direct negative
effect of species richness on connectance both using data and model simulations
(resp. $r_\delta =$ `r rdsp$ct$r` and `r rdm$ct$r`, Fig. \@ref(fig:std)) and on
average trophic level (resp. $r_\delta =$ `r rdsp$tlvl$r` and `r rdm$tlvl$r`).
Despite indirect positive effect of species richness through community biomass,
species richness had strong total negative effects on connectance in empirical
data and simulations (resp. $r_\delta =$ `r rtsp$ct$r` and `r rtm$ct$r` ) and
average trophic level (resp. $r_\delta =$ `r rtsp$tlvl$r` and `r rtm$tlvl$r`).

We performed extinction simulations scenario on the metaweb, the food-web
including all the species and nodes found in the empirical data, to better
understand how species richness change affect food-web structure. When species
are removed in a sequential order beginning by the highest trophic levels,
connectance and average trophic level decreased as we removed species (red
lines, Fig. S2a). At the opposite, removing first low trophic level species first
increase connectance and average trophic level (green lines, Fig. S2a). In turn,
random species extinction sequence had little effect on food-web structure
(black lines, Fig. S2a).

@@

@@

We found remarkable heterogeneity in community biomass and species richness
changes over the last decades in riverine fish communities, linked to changes in
food-web structure. Community biomass changes spanned from `r grbmn`% to `r
grbmx`% per decade, species richness from `r grrhn`% to +`r grrhx`% per decade,
in line previous reports of the heterogeneity of local biodiversity changes in
terrestrial, marine and freshwater realms[@dornelas_assemblage_2014;
@blowes_geography_2019; @danet_past_2024]. In turn, connectance temporal changes
spanned from `r grctn` to +`r grctx` per decade and average trophic level, from
`r grtln` to +`r grtlx` per decade. We questioned the effects of community
biomass and species richness changes over time on temporal changes in food-web
structure, that we also tested across space and with simulations using a
bioenergetic food-web model. We found positive effects of species richness
changes on changes in community biomass that were consistent across temporal and
spatial analysis and with the theoretical model. The effects of community
biomass changes on food-web structure were also consistent across analysis,
suggesting that loss of community biomass are linked to the loss of the trophic
species of high trophic level. However, changes in species richness had
contrasting effects on food-web structure depending on the type of analysis,
temporal or spatial and theoretical simulations. It suggests that food-web
structure changes induced by changes in species richness differs from the one
induced by the changes in community biomass, and that they imply different
processes across time and space. In what follows, we discuss how changes
in community biomass and species richness changes over time and space are
related to food-web assembly and disassembly processes.

We found that species richness scaled positively with community biomass, both in
temporal, spatial and theoretical simulations, in accordance with previous
analysis across space in stream communities [@woods_testing_2020;
@danet_species_2021] but also in grasslands, salt marches and forests
[@hector_plant_1999; @tilman_biodiversity_2014; @wu_relationship_2015;
@li_relationship_2018; but see @grace_integrative_2016]. This results is in
agreement with previous findings showing that food-web diversity can increase
biomass by stimulating primary production and resource uptake
[@wang_biodiversity_2018]. Our results suggest that food-web losing species
richness over time also experience a loss of biomass, while those gaining
species experience a increase in community biomass. This results is coherent the
Biodiversity-Ecosystem Functioning theory [@loreau_biodiversity_2001;
@wang_biodiversity_2018], and suggest that the loss of species richness and
biomass might impair ecosystem functioning.

Our results suggest that food-web that have undergone a biomass decline
experienced a loss of top trophic levels species and connectance, those gaining
biomass experienced a gain in top trophic levels and connectance. Importantly,
the same results were found using spatial comparison and theoretical models.
This results can be linked with the energetic and trophic biogeography theories
[@lindeman_trophic-dynamic_1942; @gravel_trophic_2011], which posits that the
presence of top predators is limited by the amount of energy and
resource available [@lindeman_trophic-dynamic_1942, @takimoto_ecosystem_2008;
@post_ecosystem_2000; @post_long_2002]. Thereby, a decrease in community biomass
can limit the amount of resource available to sustain top predators, while
increase in community biomass may result in more resource available for top
predators [@takimoto_ecosystem_2008; @post_ecosystem_2000].

Species richness temporal changes were positively associated to changes in average
trophic level, such as food-webs that have suffered a decline in species richness
have experienced a concomitant decline in average trophic level, those have seen
an increase in species richness have experienced an increase in average trophic
level. However, the effect of species richness temporal trends on average
trophic level were twice lower than the one of community biomass, and had no
effects on connectance temporal changes. Altogether, it suggests that species
richness increase and decrease were less clearly linked to specific changes in
food-web structure. Species richness temporal trends could be more linked to
changes in the redundancy of trophic roles in the food-webs, such as it has
little effects of food-web structure but increases its community biomass. This
interpretation align with the insurance hypothesis [@yachi_biodiversity_1999;
@loreau_biodiversity_2021], such as species richness and associated redundancy
in species roles can increase ecosystem stability by buffering the impacts of
perturbations [@sanders_trophic_2018].

When considering spatial analysis and model simulations, species richness
effects on food-web structure were highly distinct from the temporal analysis.
Spatial analysis and theoretical models suggest that increase in species
richness results in lower connectance and average trophic level, a well known
patterns in empirical food-web [@dunne_network_2006; @danet_species_2021]. This
result is also in line with other studies which found a mismatch between spatial
and temporal trends [@adler_contrasting_2007; @johnson_testing_2008]. This
discrepancy in space and time is due to the strong assumption that variations in
space reflect differences in assembly stages of ecological communities and then
mirror their temporal assembly/disassembly trajectories. Temporal changes in
species richness might relate to short-term dynamics while spatial gradients
reflect aggregated dynamics over a longer time scale including a far number of
processes including high species turnover along spatial gradients. Our results
then suggest community biomass changes in space might be a good surrogate to
infer assembly and disassembly of food-web in time, but that it is probably not
the case for species richness.

*Final message*

It suggests that the temporal assembly and
disassembly of food-webs are more sensitive to variations in community biomass
than in species richness, in other words that food-web structure is more robust
to changes in species richness than in biomass over time. Our study then
suggests that the changes of food-web structure over time related to changes in
species richness could be actually mediated by changes in community biomass.

@@

We investigated how food-web assemble and disassemble according to changes in
community biomass and species richness. Our study shows that temporal changes in
community biomass induced changes in food-web structure, such as decline in
biomass translates in food-webs with fewer trophic links and top predators while
increases in biomass results in food-webs with more trophic links and top
predators. Species richness temporal trends translate in low changes in food-web
structure, suggesting that species richness changes relates more to trophic role
redundancy than in changes in food-web structure. Using replicating the analysis
over spatial gradient and with a theoretical model, our study shows that
community biomass changes over space and theoretical model results in the same
changes in food-web structure than in temporal analysis, which is not the case
for species richness. It suggests that using space-for-time substitution to
infer assembly and disassembly processes in food-web structure works well for
community biomass and not species richness.

In the context of biodiversity crisis, our study reveals that declines of
community biomass results in food-web downgrading. As species of top trophic
levels plays a great role in ecosystem functioning [@estes_trophic_2011;
@wang_biodiversity_2018; @thebault_food-web_2003], it raises concerns about the
functioning of such degraded ecosystems. Loss of species richness in turn did
not affect significantly food-web structure, suggesting that it implies loss of
redundancy in trophic roles. As redundancy has a key role in buffering ecosystem
variability against perturbations [@sanders_trophic_2018;
@yachi_biodiversity_1999], it raises concerns about the stability of such
food-webs. We used here empirical freshwater food-web inferred from in-situ fish
bodysize and diet [@danet_species_2021; @bonnaffe_comparison_2021], which do not
account for the variability in invertebrates and primary producers presences,
and do not a number of processes such as changes in interaction strengths.
However, the theoretical model embedding complete food-webs showed remarkably
similar results to the empirical data.

Our study opens stimulating perspectives to further understand of biodiversity
changes alters community structure and ecosystem functioning. It is particularly
needed for ecological networks, whose structure has tremendous effects on
ecosystem functioning and stability. It implies to further investigate which
processes drive the differences in the effects of species richness
changes on food-web structure in space and time.
